# Introduction to R and RStudio

## What is R?

R is a**programming language and software environment**designed primarily for**statistical computing, data analysis, and visualization**. It is widely used in academia, research institutions, government organizations, and industry for tasks ranging from simple data summaries to advanced machine learning and reproducible research.

Key characteristics of R include:

-   **Open-source and free**: Anyone can use, modify, and distribute R.

-   **Strong statistical foundation**: Built by statisticians for statistical work.

-   **Extensive package ecosystem**: Thousands of user-contributed packages extend R’s capabilities.

-   **Excellent graphics**: R is known for high-quality, publication-ready visualizations.

-   **Reproducibility**: Scripts and documents ensure analyses can be repeated and verified.

In this workshop, R will be used as a**tool for thinking with data**, not just a calculator. You will learn how to write clear, reusable code that performs data analysis systematically.

------------------------------------------------------------------------

## Why Use R for Data Analysis?

R is especially well-suited for data analysis because it combines**data manipulation, statistical modeling, and visualization**in one environment.

### Advantages of R

-   Handles**small to very large datasets**efficiently

-   Supports**classical statistics, modern data science, and machine learning**

-   Encourages**script-based analysis**, reducing manual errors

-   Integrates well with reports (PDF, Word, HTML) through**Quarto and R Markdown**

-   Strong community support and documentation

### Typical Use Cases

-   Exploratory Data Analysis (EDA)

-   Statistical inference and hypothesis testing

-   Regression and multivariate analysis

-   Time series and forecasting

-   Data visualization and dashboards

-   Academic research and thesis work

------------------------------------------------------------------------

## Installing R

R itself is the**core engine**that performs computations.

### Steps to Install R

1.  Visit the official Comprehensive R Archive Network (CRAN):[https://cran.r-project.org](https://cran.r-project.org/)

2.  Choose your operating system (Windows / macOS / Linux)

3.  Download and install the latest stable version

After installation, R can be run from the system console. However, working directly in the R console is not ideal for beginners or large projects. This is where**RStudio**becomes essential.

------------------------------------------------------------------------

## What is RStudio?

RStudio is an**Integrated Development Environment (IDE)**for R. It makes working with R easier, faster, and more organized.

RStudio does**not replace R**. Instead:

> R does the computation, and RStudio provides a user-friendly interface to work with R.

### Why Use RStudio?

-   Code editor with syntax highlighting

-   Easy file and project management

-   Built-in help and documentation

-   Integrated plotting and visualization

-   Seamless support for Quarto documents

------------------------------------------------------------------------

## Understanding the RStudio Interface

When you open RStudio, you typically see**four panes**:

### 1. Script Editor (Top-Left)

-   Where you write and save R code

-   Supports`.R`scripts and`.qmd`Quarto files

-   Code is written once and executed many times

### 2. Console (Bottom-Left)

-   Where R executes commands

-   Displays output, messages, and errors

-   Useful for quick calculations and testing code

### 3. Environment / History (Top-Right)

-   Shows objects currently in memory

-   Displays datasets, variables, and functions

-   Helps track what data is loaded

### 4. Files / Plots / Help / Viewer (Bottom-Right)

-   File browser for project files

-   Plot display window

-   Help documentation for functions

-   Viewer for Quarto and HTML outputs

Understanding these panes is crucial for efficient workflow in R.

------------------------------------------------------------------------

## Your First Interaction with R

Let us start with very simple commands.

### Using R as a Calculator

```{r}
2 + 3
10 / 2
5^2
```

R immediately evaluates expressions and returns results.

### Assigning Values to Objects

```{r}
x <- 10
y <- 5
x + y
```

Here:

-   `<-`is the**assignment operator**

-   `x`and`y`are objects stored in memory

Object-based thinking is fundamental in R.

------------------------------------------------------------------------

## Basic Data Types in R

R works with different types of data. The most common ones are:

-   **Numeric**: Numbers (e.g., 10, 3.5)

-   **Character**: Text (e.g., "Statistics")

-   **Logical**: TRUE or FALSE

Examples:

```{r}
age <- 21
name <- "Anita"
passed <- TRUE
```

You can check the type of an object using:

```{r}
class(age)
class(name)
class(passed)
```

------------------------------------------------------------------------

## Vectors: The Core Data Structure

A vector is a**collection of values of the same type**.

```{r}
marks <- c(78, 85, 90, 88)
names <- c("A", "B", "C")
```

Key points:

-   `c()`stands for*combine*

-   All elements in a vector must be of the same type

-   Vectors are the building blocks of most R data structures

------------------------------------------------------------------------

## Working Directory and Projects

The**working directory**is the default location where R reads and saves files.

``` r
getwd()
setwd("path/to/your/folder")
```

However, changing directories manually is discouraged.

### RStudio Projects (Recommended)

-   Create a project for each workshop, assignment, or research work

-   Keeps scripts, data, and outputs organized

-   Ensures reproducibility across systems

------------------------------------------------------------------------

## Introduction to Quarto

Quarto is a**next-generation scientific and technical publishing system**that allows you to combine:

-   R code

-   Text explanation

-   Tables and figures

into a**single, reproducible document**.

### Why Quarto for This Workshop?

-   Generates**PDF, HTML, and Word**outputs

-   Ideal for teaching, reports, and books

-   Supports executable R code chunks

Example of an R code chunk in Quarto:

summary(c(10, 20, 30, 40))

------------------------------------------------------------------------

## How This Book Is Structured

This Quarto book is designed for**beginners**and progresses gradually:

1.  Basics of R and RStudio

2.  Data structures and data import

3.  Data manipulation and visualization

4.  Statistical analysis

5.  Reproducible reporting with Quarto

Each chapter includes:

-   Clear explanations

-   Annotated examples

-   Practice exercises

------------------------------------------------------------------------

## Best Practices for Learning R

-   Type the code yourself instead of copy–paste

-   Read error messages carefully

-   Experiment with small examples

-   Save scripts regularly

-   Ask*why*a result occurs, not just*what*it is

Learning R is not about memorizing commands, but about developing**data analysis thinking**.

------------------------------------------------------------------------

## Getting Help

R has extensive built-in help and a supportive community. Here are some ways to get help: - **Help files**: Use `?function_name` or `help(function_name)` to access documentation for any function. For example:

```{r}
?mean
```

-   **RStudio Help Pane**: Use the Help pane in RStudio to search for functions and topics.
-   **Online Resources**: Websites like Stack Overflow, R-bloggers, and the RStudio Community are great places to ask questions and find solutions.
-   **Books and Tutorials**: There are many free and paid resources available for learning R. Remember, seeking help is a normal part of learning programming!
