# R Environment, Syntax, and Operators

## Introduction

Before performing any data analysis, it is essential to ensure that the R environment is correctly installed and functioning. This chapter introduces the basic diagnostics for the R environment, followed by a detailed discussion of R syntax, assignment operators, variable naming conventions, and arithmetic operations. These concepts form the foundation for writing clear and correct R programs.

------------------------------------------------------------------------

## R Environment Diagnostic Checks

A properly configured R environment ensures smooth execution of scripts and reproducible results.

### Checking the R Version

```{r}
R.version.string
```

This command displays the installed R version. Using an up-to-date version of R is recommended for compatibility with modern packages.

### Testing Package Installation

```{r}
install.packages("ggplot2")
```

Successful installation confirms that R can access CRAN repositories and install external packages.

### Verifying Basic Computation

```{r}
(50 + 50) / 2
```

This simple arithmetic check confirms that the R interpreter is functioning correctly.

### Checking TinyTeX Installation

```{r}
tinytex::is_tinytex()
```

TinyTeX is required for generating PDF outputs from Quarto documents. This command verifies its availability.

------------------------------------------------------------------------

## Assignment Operators in R

R provides multiple ways to assign values to objects.

```{r}
x <- 5
y = 10
11 -> z
```

-   `<-`is the**recommended assignment operator**in R

-   `=`is commonly used but should be avoided in complex expressions

-   `->`assigns values from right to left

### Global Assignment Operator

```{r}
z <<- 15
```

The`<<-`operator assigns a value to a variable in the global environment. Its use should be limited, as it can make code harder to debug.

------------------------------------------------------------------------

## Variable Naming Conventions

Valid variable names in R:

```{r}
var1 <- 10
var_2 <- 20
var.3 <- 30
Var4 <- 40
varFive <- 50
```

Key rules:

-   Variable names are**case-sensitive**

-   They may contain letters, numbers, dots (`.`), and underscores (`_`)

-   They must not begin with a number

Invalid variable names contain spaces or special characters and will produce errors.

------------------------------------------------------------------------

## Fundamental R Syntax

R syntax is flexible and generally ignores whitespace.

```{r}
x <- 10
y <- 20
z <- x + y

a = 5; b = 15; c = a * b
d=x-y
e = x * y; f = x / y
```

Although whitespace is optional,**consistent formatting**improves readability and maintainability.

------------------------------------------------------------------------

## Basic Arithmetic Operations

R supports all standard arithmetic operations.

```{r}
5 + 3      # Addition
10 - 4     # Subtraction
6 * 7      # Multiplication
20 / 5     # Division
2 ^ 3      # Exponentiation
10 %% 3    # Modulus
10 %/% 3   # Integer division
```

These operations are frequently used in data transformation and statistical computation.

------------------------------------------------------------------------

## Summary

In this chapter, you learned how to:

-   Verify and diagnose the R environment

-   Use different assignment operators

-   Apply valid variable naming conventions

-   Understand fundamental R syntax

-   Perform arithmetic operations in R

These basics are essential for writing correct and efficient R programs. The next chapter explores**data types and type conversion**, which are crucial for understanding how R stores and processes data.
